Performance of Individual Audit Filters in Predicting Opportunities for Improvement

### 2024-06-17 manuscript.Rmd
update

---
title: "Performance of Individual Audit Filters in Predicting Opportunities for Improvement
  in Adult Trauma Patients"
subtitle: A single-center registry-based retrospective cohort
output: html_document
author: "Josef Al-Khalili"
bibliography: bibliography.bib
csl: vancouver.csl
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Abstract

@@

To reduce trauma burden, quality improvement (QI) programs have been
established by the World Health Organization (WHO) and the International
Association for Trauma Surgery and Intensive Care (IATSIC)
[@world2009guidelines]. These programs form the core of current trauma
care systems and their development around the world
[@American2022resources]. QI programs consist of a series of recommended
techniques aiming to facilitate discussion and analysis of the trauma
process in order to take corrective action for future success. The
following techniques are commonly used: a multidisciplinary mortality
and morbidity conference (MMC), a preventable death review panel, and
the use of AFs also known as quality indicators [@world2009guidelines].

Trauma care is a multidisciplinary, complex and time-critical health
care service provided by hospitals, often trauma centres.
[@American2022resources]. Trauma centres have a significantly lower
mortality rate compared to non-trauma centres, and the maturity of a
trauma centre correlates with lower mortality, demonstrating the
importance of high-quality, specialised trauma care
[@alharbi2021effectiveness; @mann1999systematic;
@candefjord2020mortality]. Despite this, even established trauma care
systems are prone to errors due to the complex nature of trauma and its
susceptibility to delays [@nikouline2021errors; @vioque2014classifying;
@o2013opportunities]. Studies between 1990 and 2014 have shown a pooled
preventable death rate of 20%, with more recent studies showing a lower
preventable death rate [@kwon2014pooled]. For example, a recent study in
a Swedish Level I trauma centre found a preventable death rate of 4%
[@ghorbani2018analysis].

### Morbidity and Mortality Conferences

The MMC is a central QI technique. It is a proven framework with the
critical components of anonymity, focusing on specific adverse events,
recognising flawed approaches, critical analysis, improvement proposals
and implementing such [@gregor2016morbidity].

The conference participants are healthcare professionals involved in
trauma care and its processes, but not necessarily the specific patient
case. The selection of patient cases varies, but common themes for
discussion are unexpected mortality, unexpected morbidity and errors
[@orlander2003morbidity].

Implemented effectively, MMCs serve as an important tool for error
reduction without being perceived as negative feedback for individual
care providers, but rather as an incentive for reporting. These
conferences promote a more comprehensive, systems-based approach that
emphasises task analysis, teamwork and quality improvement
[@lazzara2022morbidity]. The introduction of MMC has been shown to
increase morbidity and mortality reporting [@kong2016analysis].

@@

AFs, also known as quality indicators are established criteria based on
standardised care. Any deviation from the AFs should raise the attention
of health care professionals with the purpose of improving the quality
of care [@American2022resources]. An AF can involve outcomes such as
"death" or criteria related to guidelines and gold standards such as
"the placement of two large bore intravenous lines within 15 minutes
from arrival to a health care facility" [@world2009guidelines].

Previous studies have shown a reduction in trauma-related mortality
after the implementation of AFs [@chadbunchachai2001comparative;
@chadbunchachai2003study]. Per contra, newer studies seem to differ. In
2009, an extensive review by Evans et al. [@evans2009audit] sought to
determine if AFs could be used in improving processes in trauma care.
They found no studies successfully providing evidence for the use of
AFs. More than 1,500 AFs have been identified in a review by Stelfox et
al. [@stelfox2010quality]. Notably, they did not find a common set of
clearly defined, evidence-based, broadly accepted AFs for evaluating the
quality of trauma care. The perceived usefulness of audit filters still
remains high [@berg2022perceived]

Opportunities for improvement (OFIs) are the endpoint of patient care.
Following the review of individual patient cases there is a decision
made regarding the existence of an OFI [@American2022resources]. It has
been shown that such review processes are associated with high-quality
trauma care [@santana2014development]. OFIs are typically associated
with failures in initial care [@vioque2014classifying], specifically in
airway management, fluid resuscitation, haemorrhage control and chest
injury management [@sanddal2011analysis; @roy2017learning;
@o2013opportunities].

There is lacking evidence for the use of AFs for improving trauma care
[@evans2009audit; @stelfox2010quality] and the focus has been on
mortality as outcome measure [@chadbunchachai2001comparative;
@chadbunchachai2003study]. Furthermore, the use of AFs has also been
associated with high frequencies of false positive OFI predictions,
ranging from 24% to 80% [@sanddal2011analysis; @roy2017learning], and
some AFs do not appear to correlate with OFIs at all [@lewis2017delay].

@@

```{=html}
<!-- Cohort study: Give the eligibility criteria, and the sources and
methods of selection of participants. Describe methods of
follow-up. For matched studies, give matching criteria and number of
exposed and unexposed -->
```
```{=html}
<!-- Case-control study: Give the eligibility criteria, and the
sources and + methods of case ascertainment and control
selection. Give the rationale for the choice of cases and
controls. For matched studies, give matching criteria and the number
of controls per case -->
```
```{=html}
<!-- Cross-sectional study: Give the eligibility criteria, and the
sources and methods of selection of participants -->
```
## Variables and data sources/measurements
The outcome variable in this study is an OFI, as established by the MMCs
and furthermore treated as a dichotomous variable with "Yes - At least
one OFI identified" and "No - No OFI identified".
The 11 AFs used at Karolinska University Hospital served as the exposure variable, as shown in Table 1. Each AF is labeled "original" or "manually created." Manually created AFs were calculated using trauma registry data, while original AFs were recorded in the quality database based on criteria beyond the available registry data.

The data spanning 2012 to 2022 was retrieved from the
Karolinska University Hospital trauma registry and trauma care quality
database. Patient data on vital signs, care processes and interventions, level of
care and time aspects was retrieved from SweTrau while both exposures
(AFs) and outcome (OFI) were retrieved from the Karolinska University
Hospital trauma care quality database. All data was anonymised to protect patient privacy.
```{=html}
<!-- Clearly define all outcomes, exposures, predictors, potential
confounders, and effect modifiers. Give diagnostic criteria, if
applicable. For each variable of interest, give sources of data and
details of methods of assessment (measurement). Describe comparability
of assessment methods if there is more than one group -->
```
## Bias
Consensus decisions at MMC conferences may introduce misclassification bias. Additionally, AF flagging by specialized nurses is a manual process, potentially leading to selection bias. Healthcare professionals attending MMCs may not accurately recall key details impacting OFI decisions, introducing recall bias. Furthermore, professionals not present for specific cases might over- or underestimate OFI significance.
<!-- Describe any efforts to address potential sources of bias -->

@@

Notably, OFI in this study was not soley a reflection
of mortality but also morbidity. Rather, it could highlight differences in quality gaps
between hospitals.

>>>>

---
title: Performance of Individual Audit Filters in Predicting Opportunities for Improvement
  in Adult Trauma Patients
author: "Josef Al-Khalili"
output:
  html_document: default
  pdf_document: default
subtitle: "A single-center registry-based retrospective cohort"
bibliography: bibliography.bib
csl: vancouver.csl
editor_options:
  markdown:
    wrap: 72
---
```{r sourcecode, echo=FALSE, fig.align="center", warning=FALSE, message=FALSE,error=FALSE}
source("main.R")
library(rofi)
library(dplyr)
library(ggplot2)
library(pROC)
library(gtsummary)
library(gt)
library(boot)
library(webshot2)
library(knitr)
```
# Abstract

@@

To reduce trauma burden, quality improvement (QI) programs have been
established by the World Health Organization (WHO) and the International
Association for Trauma Surgery and Intensive Care (IATSIC)
[@world2009guidelines]. These programs form the core of current trauma
care systems
[@American2022resources] and commonly include multidisciplinary mortality
and morbidity conferences (MMC), preventable death review panels, and
the use of AFs also known as quality indicators [@world2009guidelines].
Despite QI programs, trauma care remains prone to errors [@nikouline2021errors; @vioque2014classifying;
@o2013opportunities]. Studies have shown a pooled preventable death rate of 20% from 1990 to 2014, with more recent studies indicating lower rates [@kwon2014pooled]. For instance, a recent study in
a Swedish Level I trauma centre found a preventable death rate of 4%
[@ghorbani2018analysis].

### Morbidity and Mortality Conferences
The MMC is a central QI technique focusing on anonymity, critical analysis of specific adverse events, recognizing flawed approaches, and proposing and implementing improvements [@gregor2016morbidity]. Attended by healthcare professionals involved in trauma care, MMCs discuss themes like unexpected mortality, morbidity, and errors [@orlander2003morbidity], fostering error reduction and encouraging reporting without negative feedback [@lazzara2022morbidity]. Effective MMC implementation has been shown to increase morbidity and mortality reporting [@kong2016analysis].

@@

AFs are criteria based on standardized care to enhance trauma care quality. Deviation from AFs should raise the attention
of health care professionals with the purpose of improving the quality
of care [@American2022resources]. AFs can involve outcomes like "death" or adherence to guidelines, such as "placement of two large bore intravenous lines within 15 minutes of arrival." [@world2009guidelines].Studies have shown mixed results on the effectiveness of AFs in reducing trauma-related mortality. While some early studies reported mortality reduction [@chadbunchachai2001comparative;
@chadbunchachai2003study], a 2009 review by Evans et al. [@evans2009audit] found no substantial evidence supporting AFs' effectiveness. Stelfox et al. [@stelfox2010quality] identified over 1,500 AFs but noted a lack of a common, evidence-based set for evaluating trauma care quality, despite the high perceived usefulness of AFs [@berg2022perceived].

Opportunities for improvement (OFIs) are decided upon review of individual patient cases [@American2022resources]. They are typically associated
with failures in initial care [@vioque2014classifying] like
airway management, fluid resuscitation, haemorrhage control and chest
injury management [@sanddal2011analysis; @roy2017learning;
@o2013opportunities]. There is lacking evidence for the use of AFs for improving trauma care
[@evans2009audit; @stelfox2010quality] and the focus has been on
mortality as outcome measure [@chadbunchachai2001comparative;
@chadbunchachai2003study]. Furthermore, the use of AFs has been
associated with high frequencies of false positives,
ranging from 24% to 80% [@sanddal2011analysis; @roy2017learning], and
some AFs do not appear to correlate with OFIs at all [@lewis2017delay].

The aim of this study is to determine the performance of AFs in
predicting OFIs and furthermore assess their relevance as a screening
tool for MMCs.

@@

```{=html}
<!-- Cohort study: Give the eligibility criteria, and the sources and
methods of selection of participants. Describe methods of
follow-up. For matched studies, give matching criteria and number of
exposed and unexposed -->
```
```{=html}
<!-- Case-control study: Give the eligibility criteria, and the
sources and + methods of case ascertainment and control
selection. Give the rationale for the choice of cases and
controls. For matched studies, give matching criteria and the number
of controls per case -->
```
```{=html}
<!-- Cross-sectional study: Give the eligibility criteria, and the
sources and methods of selection of participants -->
```
## Variables and data sources/measurements
The outcome variable is an OFI, as established by the MMCs
and treated as a dichotomous variable with "Yes - At least
one OFI identified" and "No - No OFI identified".
The 11 AFs used at Karolinska University Hospital served as the exposure variable, as shown in Table 1. Each AF is labeled "original" or "manually created." Manually created AFs were calculated using trauma registry data, while original AFs were recorded in the quality database based on criteria beyond the available registry data.

The data spanning 2012 to 2022 was retrieved from the
Karolinska University Hospital trauma registry and trauma care quality
database. Patient data on vital signs, care processes and interventions, level of
care and time aspects was retrieved from SweTrau while both exposures
(AFs) and outcome (OFI) were retrieved from the Karolinska University
Hospital trauma care quality database. All data was anonymised to protect patient privacy.
```{=html}
<!-- Clearly define all outcomes, exposures, predictors, potential
confounders, and effect modifiers. Give diagnostic criteria, if
applicable. For each variable of interest, give sources of data and
details of methods of assessment (measurement). Describe comparability
of assessment methods if there is more than one group -->
```
## Bias
Consensus decisions at MMC conferences may introduce misclassification bias. Additionally, AF flagging by specialized nurses is a manual process, potentially leading to selection bias. Healthcare professionals attending MMCs may not accurately recall key details impacting OFI decisions, introducing recall bias. Furthermore, professionals not present for specific cases might over- or underestimate OFI significance.

@@

Notably, OFI in this study was not soley a reflection
of mortality but also morbidity. Rather, it could highlight differences in quality gaps
between hospitals.
Trauma care is a multidisciplinary, complex and time-critical health
care service provided by hospitals, often trauma centres.
[@American2022resources]. Trauma centres have a significantly lower
mortality rate compared to non-trauma centres, and the maturity of a
trauma centre correlates with lower mortality, demonstrating the
importance of high-quality, specialised trauma care
[@alharbi2021effectiveness; @mann1999systematic;
@candefjord2020mortality]. 
 It has
been shown that such review processes are associated with high-quality
trauma care [@santana2014development].
